Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2020-08-27 write-up/covid-writeup.tex
Update covid-writeup.tex

This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling.

@@

\subsection{Imperfect testing}

@@

\section{Doubly robust estimation}

The prior section focused on the negative consequences.

The question is "What can be done".

Here we present a technique for combining epidemilogical forecasts, IPW built using auxiliary information, and measurement-error corrections to generate 'doubly robust' estimates of the prevalence (true case count).

We provide an appropriate extension of the statistical decomposition.
Remark that the robustness is under certain assumptions which we know are likely false; therefore, we are interested in

@@

With non-probability samples, one can often reduce bias in two ways.  First, one may model the propensity for self-selection and use inverse probability weighting (IPW)~(cite: Elliott) techniques to adjust for selection bias.  In causal inference~(cite: Robins), given certain ignorability assumptions, the observational data permits direct estimation of treatment propensity.  In the COVID-19 context, one does not observe those who decide not to be tested.  In order to estimate selection propensities, one instead needs auxiliary information.  Here we present two methods for propensity estimation -- the first leveraging population-level summary statistics and the second leveraging the probability sample.

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde R_i$ be an indicator that the individual was included in the probability sample and $\tilde W_i$ be the probability of inclusion for the $i$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
$$
\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
$$
Taking the expecation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparadness district)

In the simple stratified case with a non-parametric selection propensity, the weight is given by estimator of the
quantity $\sum_k N \cdot \tilde d_k/n \log ( 1- \pi_k )$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

\subsubsection{An IPW estimator}

Given the

Then our new estimator is
$$
\bar y_n^\star
= \frac{1}{1-FP-FN} \cdot \frac{1}{N} \sum_{i=1}^n \frac{y_i - FP}{\pi (x_i; \hat \theta)}
= \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k}{N \pi_k} (\bar y_k - FP).
$$
where $k$ indexes the strata and $d_k$ is the number of samples in strata $k$.

\subsection{Statistical decomposition of error}

For $i=1,\ldots,N$ let $\tilde Y_i = (Y_i - FP)/ (1- FP -FN)$ and $\tilde I_i = I_i / \pi_{K_i}$.  Then the error can now be expressed as:
$$
\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \sigma_{Y} \times D_M
$$

@@

We circle back and add in a SEIR model with measurement-error.  We can correct for selection bias two through simple rules (see Song and Mukherjee and others for more complex alternatives).  With such limited data we can only predict total counts and use the demographics as is to be representative.  This yields estimates at $\hat y_k$

@@

@@

@@

@@

- Easy access to history of testing requirements
- Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
- Record symptom/asymptomatic status!

>>>>

This article discusses the relationship between three statistical concepts: selection bias, measurement error, and the too often forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical arguments, we clarify five important issues related to COVID-19 case-count analysis.  First, we show that unadjusted prevalence rates are (unsurprisingly) biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates includes an additional term in the error decomposition that captures the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts overestimates the true rate of change prior to the peak time and underestimates it immediately after.  This implies a data scientist analyzing observed rates will be (on average) overly pessimistic in the early stages of the pandemic, and overly optimistic subsequent to the peak times.  We show that estimates of the effective reproduction number may be similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a sensitivity analysis method for estimating data quality and its use in understanding the potential biases in COVID-19 case-count analyses.  {\bf We then end the paper with a discussion of the benefits of randomized testing and use of auxiliary information to improve over simple random sampling. NEED TO MAKE CLEAR THAT WE CAN MAKE USE OF RANDOM SAMPLING TO IMPROVE INFERENCE FROM NONPROB SAMPLES}

@@

\subsection{Imperfect testing}
\label{section:imperfecttesting}

@@

\section{Potential improvements to prevalence estimation}

The prior section presented negative consequences of analysing the measurement-error adjusted prevalence from observed case-count data.  This raises the immediate question as to whether prevalence estimators can be improved using publicly available data.  In this section, we consider two directions. First, we use auxiliary information to build selection propensities and consider an inverse probability weighting (IWP) estimator.
We extend the statistical error decomposition to the weighted setting.
Second, we consider an extension of compartmental models from infectious disease epidemiology that adjust for selection bias and measurement-error as model-based forecasts of prevalence.
We then combine these epidemiological forecasts, selection propensities, and measurement-error corrections to generate \emph{doubly robust} prevalence estimators. We provide an appropriate extension of the statistical error decomposition and compare to the other prior decompositions. The proposed method and corresponding statistical error decompositions guide our recommendations in Section XX.

@@

With non-probability samples, one can often reduce bias in two ways.  First, one may model the propensity for self-selection and use inverse probability weighting (IPW)~(cite: Elliott) techniques to adjust for selection bias.  In causal inference~(cite: Robins), given certain ignorability assumptions, the observational data permits direct estimation of treatment propensity.  In the COVID-19 context, one does not observe those who decide not to be tested.  In order to estimate selection propensities, one instead needs auxiliary information.  Here we present two methods for propensity estimation -- use of population-level summary statistics and use of a probability sample.

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde R_i$ be an indicator that the individual was included in the probability sample and $\tilde W_i$ be the probability of inclusion for the $i$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
$$
\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
$$
Taking the expecation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparadness district).  Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

\subsubsection{An IPW estimator and statistical error decomposition.}
Given a selection propensity, define a weight~$W(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
$$
\bar y_n^\star
= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\pi (x_i; \hat \theta)}
\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k}{N \pi_k} (\bar y_k - FP),
$$
where the second equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata and $d_k$ is the number of samples in strata $k$. Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.

Let $\tilde I_j = I_j / (N \cdot \pi (X_j; \theta)) := I_j W_j$ for $j=1,\ldots,N$.  Then we can express $\bar y_n^\star$ as a weighted estimator $(1-FP-FN)^{-1} \sum_{i=1}^n $Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
$$
\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M} \times \frac{E_J[\tilde I_J]}{N}
$$
where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$.  See Appendix XX for the derivation.

Interestingly, the impact of measurement-error on data quality is unchanged when considering a weighted estimand.  Similar to~\cite{Meng2018}, the data quality index....

@@

Until now, this paper focused primarily on selection bias in coronavirus case-count data from a purely statistical perspective.  Here, we extend the approach to whether it is possible bring in simple models from infectious diease epidemiology.  Our goal is simply an estimate of $\bar y_k$, the prevalence in strata $k$.

We rely heavily on recently available software~\cite{Song2020} for estimation of such models.

We circle back and add in a SEIR model with measurement-error.  We can correct for selection bias two through simple rules (see Song and Mukherjee and others for more complex alternatives).  With such limited data we can only predict total counts and use the demographics as is to be representative.  This yields estimates at $\hat y_k$ for each strata.

@@

Corresponding standard errors are
$$
\text{Var} (\bar y_{DR}^\star) =
$$
See Appendix XX for a derivation.

{\bf where to describe difference from Chen}.

Remark that the robustness is under certain assumptions which we know are likely false; therefore, we are interested in

@@

\subsection{COVID-19 active infection prevalence in Indiana}

We have a small non-random sample to supplement and get these; then use raking on the larger study.  Appendix XX presents the marginal distributions for the non-random sample per

Heuristically, one may suppose that asymptomatic individuals are less likely to select.  This is born out.

\begin{rmk}[Use of second nonrandom sample]
The key issue with selection bias is whether the individual is displaying symptoms and had contact with someone who was COVID+.  These are not captured by the

In random sample, $44.2$\% of individuals who tested positive for SARS-CoV-2 positive individuals reported no symptoms during the 2 weeks before testing.

In nonrandom sample, $20.2$\% of individuals who tested positive for SARS-CoV-2 reported no symptoms during the 2 weeks before testing.

\end{rmk}

@@

SEROPREVALENCE (i.e., used to be infected).  We can attempt to estimate using our methods and assume non-overlapping samples.  Then back out the correlation.

@@

- Easy access to history of testing requirements
- Smaller random samples over time (at least as testing and prevalence changes drastically) to assess selection changes
- Record symptom/asymptomatic status and COVID contact.  If not for all, then for at least some we need these covariates!!!!!
